Testing the motor and cognitive foundations of Paleolithic social transmission

### 2021-08-16 manuscript/manuscript.Rmd
add some toy figures and experiment with figure crossreferencing.

---
title: Testing the motor and cognitive foundations of Paleolithic social transmission
author: 
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
- Megan Beney Kilgore^[Department of Anthropology, Emory University, Atlanta, GA, USA; megan.elizabeth.beney@emory.edu]
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
abstract: "Stone tools provide key evidence of human cognitive evolution but remain difficult to interpret. Toolmaking skill-learning in particular has been understudied even though: 1) the most salient cognitive demands of toolmaking should occur during learning, and 2) variation in learning aptitude would have provided the raw material for any past selection acting on tool making ability. However, we actually know very little about the cognitive prerequisites of learning under different information transmission conditions that may have prevailed during the Paleolithic. This paper presents results from a pilot experimental study to trial new experimental methods for studying the effect of learning conditions and individual differences on Oldowan flake-tool making skill acquisition. We trained 23 participants for 2 hours to make stone flakes under two different instructional conditions (observation only vs. direct active teaching) employing appropriate raw materials, practice time, and real human interaction. Participant performance was evaluated through analysis of the stone artifacts produced. Performance was compared both across experimental groups and with respect to individual participant differences in grip strength, motor accuracy, and cognitive function measured for the study. Our results show aptitude to be associated with fluid intelligence in a verbally instructed group and with a tendency to use social information in an observation-only group. These results have implications for debates surrounding the cumulative nature of human culture, the relative contributions of knowledge and know-how for stone tool making, and the role of evolved psychological mechanisms in “high fidelity” transmission of information, particularly through imitation and teaching.\\par \\textbf{Keywords:} Oldowan; Stone toolmaking; Social learning; Individual variation; Cognitive aptitudes; Motor skills"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

“*A creature that learns to make tools to a complex pre-existing pattern...must have the kind of abstracting mind that would be of high selective value in facilitating the development of the ability to communicate such skills by the necessary verbal acts.*”[@montagu1976: 267]

@@

We selected a two-hour learning period for both pragmatic and theoretical reasons. Pargeter et al. [@pargeter2019] found that even \~90 hours of fully interactive instruction and practice was insufficient to achieve handaxe-making skills comparable to the later Acheulean site of Boxgrove [@garcía-medrano2019; @stout2014], and estimated actual time to mastery as ranging from 121 to 441 hours for different participants. However, they observed the greatest, fastest, and most individually variable skill increases during the first 20 hours of practice. In addition, initial performance was moderately correlated with later achievement. This suggests that studying early-stage learning may be a pragmatic alternative, especially for research investigating individual differences in aptitude. Studies of simple flake production similarly document large initial variation [@stout2015] and rapid early progress [@stout2015; @putt2019; @stoutKnappingSkillEarliest2006]. We designed the current study to test the utility of studying learning and variation during the first two hours of simple flaking instruction/practice, in hopes of finding a viable compromise between experimental realism and cost

@@

@@

\

@@

## Stone Tool Making
After individual difference testing, participants engaged in a 2-hour stone tool making session, with a 15-minute break after 1 hour. Participants were instructed not to seek out additional training or information on stone tool making (i.e., via the internet) during these breaks. Each group of participants was randomly assigned to one of two experimental conditions: no teaching or teaching. In both conditions, participants were first given an opportunity to inspect and handle examples (**Figure**) of the kind of stone tools (flakes) they are being asked to produce. They were told that their objective was to produce as many flakes as possible from the materials provided. This meant that even the untaught condition included some minimal instruction (being told the objective) , however this was considered to be unavoidable without creating a much more elaborate and naturalistic context in which participants would develop their own technological goals. Such a design would also be expected to increase behavioral variability, demanding correspondingly larger samples of participants to identify patterns and making direct comparisons with the taught condition.

### Raw Materials
Each participant was provided with 9 cores for use over the 2-hour experiment. These cores were produced from larger chunks of a fine-grained basalt purchased from neolithics.com by fracturing them with a sledgehammer. This produced irregular, angular chunks for use in the experiment, weighing between 459g - 1876g (mean = 975g). All cores were weighed, measured (Length, Width, Thickness), and painted white so that new fracture surfaces could be discriminated from those created during production. Cores were sorted by shape and weight and then distributed evenly to each participant. As a result, there were no significant difference across participants in the mean weight (ANOVA, df = 22, F=0.3, p = 0.9; Levene test of homogeneity of variance = 1.04, df1=22, df2 = 184, p = 0.4) or shape (Length $\times$ Width/Thickness: ANOVA, df = 22, F=0.4, p = 0.9; Levene statistic = .6, df1=22, df2 = 184, p = 0.9) of cores provided. This was also true comparing the two experimental conditions (Taught vs. Untaught mean weight = 1001g vs. 956g, t = 1.24, df = 205, p = 0.2, Levene's Test F = 0.6, p = 0.4; mean shape = 221.43 vs. 221.45, t = -0.003, df = 205, p = 0.9, Levene's Test F = 3.8, p = 0.05). Participants were, however, allowed to choose which cores to work on so that differences in the weight and shape of cores actually used across participants and conditions could still emerge as a result of selection bias.

Sixty pounds of 3-to-5 inch basalt "Mexican Beach Pebbles" were purchased from a landscaping supply company for use as hammerstones in the experiment. Of these, 90 were selected as suitable for use. These weighed between 213g-1360g (mean = 425) and varied in elongation (L/W = 1.01 to 2.65) and relative thickness (LxW/T = 90.48 to 283.67). Forty-five stones were placed in the middle of the knapping area (Figure 2) for participants to freely choose from during the experiment. Broken hammerstones were replaced from the reserve to maintain a consistent number and range of choices. Each hammerstone was numbered and participants' choices were recorded along with the number of the core(s) being worked on with a particular hammerstone.\

### Experimental Conditions
In both conditions, three researchers were present to record activities and collect materials. Participants were seated in a circle [@shea2015] and experiments were video recorded using two cameras (Figure 1). Participants were free to select hammerstones from the common pile and to work on any or all of their nine assigned cores in any order they preferred. However, each core and all associated debitage were collected before participants were allowed to start working on a new core, so it was not possible to partially work and then return to a particular core later. The order of cores used and associated hammerstones were recorded for each participant during the experiment.

@@

## Lithic Analysis
All finished cores were weighed and measured (L, W, T). Delta weight was calculated as (Start weight-End weight)/Start weight. All detached pieces (DPs) were collected and weighed. We did not sort DPs into types (e.g., whole flakes, fragments) as this would have greatly increased processing time and it is not clear that such distinctions add relevant information regarding utility/desirability beyond that supplied by metrics [@stout2019]. All DPs larger than 40mm in maximum dimension were photographed and measured. It is conventional in Early Stone Age lithic analysis to employ a 20 mm cut-off. We selected a higher threshold for both pragmatic (analysis time) and theoretical reasons. Flake use experiments have shown that flakes weighing less than 5--10 g or with a surface area below 7--10 cm2 [@prasciunas2007] or with a maximum dimension \<50-60 mm [@key2014] become markedly inefficient for basic cutting tasks. Similarly, data from Oldowan replication experiments [@stout2019] show that the utility index (flake cutting edge/flake mass1/3) \* (1 - exp[-0.31 \* (flake maximum dimension -- 1.81)]) developed by Morgan et al. (2015) falls off rapidly below 40mm maximum dimension ( Mean Utility \< 40mm = 0.508; \>=40mm = 0.946; t= 11.99, df = 707, p \< 0.000). By including weight in our cut-off criteria we also avoid skewing the flake shape distribution by selectively retaining long, thin pieces (i.e., MD \> 40, weight \< 5g) while discarding rounder pieces of similar (or greater) weight and area.

@@

## Statistical Analyses
To evaluate the association between psychometric, motor-skill, and training measures and technological outcomes, we adopted an information-theoretic approach [@burnham2002]. Information-theoretic approaches provide methods for model selection using all possible combinations of variables while avoiding problems associated with significance-threshold stepwise selection. We used the corrected Akaike information criterion (AICc) to rate each possible combination of predictors on the balance between goodness of fit (likelihood of the data given the model) and parsimony (number of parameters). The AICc consists of the log likelihood (i.e., how well does the model fit the data?) and a penalty term for the number of parameters that must be estimated in the model (i.e., how parsimonious is the model?), with a correction for small sample sizes (AICc converges to the standard AIC at large samples). A lower AICc indicates a more generalizable model and we used it to compare and rank various possible models. Each analysis begins with a full model that includes all predictors of interest. All possible combinations of predictors are then fit, and the resulting models are ranked and weighted based on their AICc. The "best" model is chosen because it has the lowest AICc score.

@@

\

# Results
Following a recent protocol to enhance the reproducibility and data transparency of archaeological research [@marwick2017], detailed results of all analyses and assessments of the data structure are available in our paper's supplementary materials and through Github (<https://github.com/Raylc/PaST-pilot>). Here we limit discussion to the major ﬁndings regarding flaking performance and individual differences. We were particularly interested in: 1) group level effects of experimental condition, 2) individual differences in aptitude and learning, and 3) potential interactions between learning conditions and individual differences. To address these questions, we employed data reduction (Principal Component Analysis) to derive two summary metrics of flaking performance, compared these factors across the two experimental conditions, and built multivariate models examining the relations between our various psychometric measures, subject’s motor skill scores, and our two lithic performance factors.

## Principal Component analyses
The following two sections outline factor analyses designed to summarize our main study metrics tracking individual variation in DP sizes and shapes and lithic performance measures.

### Detached Piece size and shape
To better understand the relationship between DP shape and training/individual variation, we entered our nine flake linear plan measurements along with maximum flake length and thickness into a principal component analysis (PCA) from which summary coordinates were extracted. Bartlett's Test of Sphericity was significant ($\chi ^2$ (10) =4480, p \< .01) indicating that the set of variables are adequately related for factor analysis.

@@

### Lithic flaking performance measures
To better understand the relationship between our various lithic performance measurements and to reduce data dimensionality, we conducted a second principal component analysis examining the study’s six lithic performance measures (count of large pieces [>40mm and 5g], mass of large pieces relative to total detached mass, core delta mass, and the three flake shape factors). All of these measures were summarized for each core and unique factor scores were calculated from these core-specific measures. Bartlett's Test of Sphericity was significant ($\chi ^2$ (6) =3185, p \< .01) indicating that the set of variables are at least adequately related for factor analysis.

The analysis yielded two factors explaining a total of 56% of the variance for the entire set of variables (Table). Factor 1 (hereafter “Quantity”) explains 28.7% of the variance and tracks flaking quantity due to high positive loadings on large DP count and mass ratio and on core delta mass. Performance factor 2  (hereafter “Quality”) covers 27% of the sample variance and measures flaking quality as reflected in high positive loadings on Shape Factors 1 (size) and 3 (thin, “flake-like” shape) and a negative loading on Shape Factor 2 (“splinter-like” thickness and elongation). High scores on Quality thus reflect production of larger, relatively thinner, and more typically flake-shaped vs. splinter-shaped DPs.

@@

## Relationships between Performance Measures
This approach also allowed us to compare the relationship between Total Productivity, Quantity, and Quality across our two experimental groups (Figures). As might be expected, we found that per-core Quantity and Total Productivity are positively correlated in both groups (Fig. Xa), although this relationship is twice as strong in the trained (F[1, 9] = 33, p \< 0.01, Adj. $R^2$ = 0.8) compared to untrained (F[1, 8] = 8, p = 0.02, Adj. $R^2$ = 0.4) group. Interestingly, we also found evidence of a negative correlation between Total Productivity and Quality in the untrained group (F[1, 8] = 28, p = \<0.01, Adj. $R^2$ = 0.7), but no relation in the Trained group (Fig. Xb). A qualitatively similar trend with respect to Quality vs. Quantity (Fig. Xc) did not achieve significance (F[1, 17] = 0.6, p = 0.2).

@@

## Do trained, untrained, and expert knappers perform differently?
Here we compare our flaking outcomes (DP size/shape and flaking performance factors) between the trained and untrained groups. Our expert demonstrator/instructor is included as a performance benchmark.

Table summarizes the results of ANOVA tests group level difference in central tendency on various performance measures. We found no significant differences between the trained and untrained groups on our flaking Quantity and Quality factors. In contrast, three-way flake size and shape comparisons between our expert knapper and the two novice groups showed that the expert knapper made significantly more large flakes (effect size = 0.14), had a significantly higher core delta mass signal than either of the novice groups (effect size = 0.26), and left significantly smaller finished cores (effect size = 0.27) (Figure). All three of these results show either medium or large effect sizes. In all three comparisons, the trained group’s data distributions tended towards the expert sample although they were not significantly different from the untrained group (Figures-Core examples too). We also observed a significant difference in shape factor 2 (splinters) driven by the expert’s lower values, but with a very low (<0.01) effect size. These results show that mean core reduction intensity and large flake production rates distinguish expert and novice performance whereas novices in experimental groups produced pieces of similar mean size and shape as those of the expert trainer.

@@

## Does performance change over time?
In addition to comparing overall performance during the two hour experiment, we also wanted to determine if groups or individuals differed in learning (i.e., performance change) over the period.  For these analyses, we calculated the learning stage as the ordinal number of each core out of the total number knapped by each subject (i.e., core 2 of 4 or 4 of 8 both equal 50% complete). These relative core use-order percentages were then binned into 20 percent brackets for core-order and group-level comparisons. Flaking outcomes were tracked using the two performance factors (Quality and Quantity). We added the nodule starting mass to track whether training/practice times impacted raw material selection.

@@

## Do individual differences in motor skill and psychometric measures predict flaking performance?
One of the experiment’s primary goals was to test if measures of individual perceptual-motor and cognitive variation predict success in stone flaking across different training conditions. To address this goal, we built three multivariate models examining the relations between training conditions, individual difference measures, and our three lithic performance measures (overall productivity and average per-core Quantity and Quality). These models enabled us to determine which of the psychometric and motor skill factors are better predictors of a participant's flaking performance in the study.

We considered all possible interactions between five individual difference measures, core size, training condition, and the three performance measures, with each subject providing one data point. Each model’s continuous predictors (highest n-back level, Raven's Progressive Matrix score, BEAST score, starting nodule mass, Fitts score, and grip strength) were centered such that zero represents the sample average, and units are standard deviations. Our two motor skill and strength measures (grip strength and Fitt's performance scores) are also strongly correlated (F [1,19] = 15, p \< 0.01, $R^2$ = 0.41). However, these two measures track complementary components of athleticism (strength vs. speed/accuracy tradeoffs) and so we decided to include both in the model selection process.

The full models were fitted with the lm function in R 3.2.3, and we used the Glmulti package’s automated model selection algorithm to select the best performing model (lowest AICc score) (see methods for further details on the multimodal selection process). All three models follow the same complete model statement as follows:

*Flaking performance variable $\sim$ Training condition + Highest n-back level + Raven’s Progressive Matrix score + BEAST score + Fitt’s score + Grip strength*

For our two per-core performance factors (Quantity and Quality) it is also relevant to consider how individual core features may have affected performance. We found no evidence of individual or group level practice effects over the two hours, so we did not include core order in the models. We did, however, find that subjects selected progressively larger nodules throughout the experiment. It is thus important to understand whether nodule variability had any impact on our flaking results. Because starting nodule size (mass) and shape were strongly correlated (F [1,157] = 186, p < 0.01, $R^2$ = 0.54) we included nodule mass as a covariate to control for any variance in flaking performance that may be driven by nodule differences.

### Model 1: Individual differences and quantity flaking
Our first model examined variance in overall flaking productivity measured by each subject’s combined flaked mass (nodule starting mass-core final mass). This provides a basic measure of variation in individuals’ success detaching pieces and reducing cores from a standardized (see Methods) raw material supply. From the same candidate pool size of 55893 possible multivariate models, the best performing model returned an AICc value of -18 (Average AIC = -13). This model comprised the following statement with two main and three interaction effects:

*Total flaked mass $\sim$ Training condition + Grip strength + RPM$\times$Highest n-back level + Fitts score$\times$BEAST score + Grip strength$\times$RPM*

This model explains a statistically significant and substantial proportion of variance in flaking productivity ($R^2$ = 0.84, F (6, 14) = 12.7, p < 0.01, adj. $R^2$ = 0.77). A model residuals normality test shows no significant differences with the normal distribution (p = 0.72) indicating that this relationship is linear. A Breusch-Pagan test showed no evidence for heteroskedasticity (BP = 2, df = 6, p = 0.8).

Table presents this model’s coefficients and summary outputs, wherein baseline refers to the untrained condition with all continuous predictors at the sample average. The parameter estimates for the continuous predictors reflect the expected change in utility for 1 standard deviation change in the predictor variable. We found significant (p< 0.05) and substantial (Standardized Estimate >= 0.50, i.e. a 50% change in variable) main effects of Grip Strength, Visuospatial nBack, and BEAST. The main effect of Grip Strength (Figure), irrespective of learning condition, indicates the basic importance of strength in generating higher production rates among naive knappers at least when efficiency and quality are not considered.

@@

### Model 2: Individual differences and quality flaking
The second full model examined the variance in average flaking Quantity per core. It thus complements our first model assessing overall productivity by testing for differences in reduction intensity at the level of individual cores. From a candidate pool of 55893 possible multivariate models, the best performing model returned an AICc value of 32 (Average AIC = 44). This model comprised the following statement with three main and four interaction effects:

>>>>

---
title: Testing the motor and cognitive foundations of Paleolithic social transmission
author: 
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
- Megan Beney Kilgore^[Department of Anthropology, Emory University, Atlanta, GA, USA; megan.elizabeth.beney@emory.edu]
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
abstract: "Stone tools provide key evidence of human cognitive evolution but remain difficult to interpret. Toolmaking skill-learning in particular has been understudied even though: 1) the most salient cognitive demands of toolmaking should occur during learning, and 2) variation in learning aptitude would have provided the raw material for any past selection acting on tool making ability. However, we actually know very little about the cognitive prerequisites of learning under different information transmission conditions that may have prevailed during the Paleolithic. This paper presents results from a pilot experimental study to trial new experimental methods for studying the effect of learning conditions and individual differences on Oldowan flake-tool making skill acquisition. We trained 23 participants for 2 hours to make stone flakes under two different instructional conditions (observation only vs. direct active teaching) employing appropriate raw materials, practice time, and real human interaction. Participant performance was evaluated through analysis of the stone artifacts produced. Performance was compared both across experimental groups and with respect to individual participant differences in grip strength, motor accuracy, and cognitive function measured for the study. Our results show aptitude to be associated with fluid intelligence in a verbally instructed group and with a tendency to use social information in an observation-only group. These results have implications for debates surrounding the cumulative nature of human culture, the relative contributions of knowledge and know-how for stone tool making, and the role of evolved psychological mechanisms in “high fidelity” transmission of information, particularly through imitation and teaching.\\par \\textbf{Keywords:} Oldowan; Stone toolmaking; Social learning; Individual variation; Cognitive aptitudes; Motor skills"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

"*A creature that learns to make tools to a complex pre-existing pattern...must have the kind of abstracting mind that would be of high selective value in facilitating the development of the ability to communicate such skills by the necessary verbal acts.*"[@montagu1976: 267]

@@

We selected a two-hour learning period for both pragmatic and theoretical reasons. Pargeter et al. [@pargeter2019] found that even \~90 hours of fully interactive instruction and practice was insufficient to achieve handaxe-making skills comparable to the later Acheulean site of Boxgrove [@garcía-medrano2019; @stout2014], and estimated actual time to mastery as ranging from 121 to 441 hours for different participants. However, they observed the greatest, fastest, and most individually variable skill increases during the first 20 hours of practice. In addition, initial performance was moderately correlated with later achievement. This suggests that studying early-stage learning may be a pragmatic alternative, especially for research investigating individual differences in aptitude. Studies of simple flake production similarly document large initial variation [@stout2015] and rapid early progress [@stout2015; @putt2019; @stoutKnappingSkillEarliest2006]. We designed the current study to test the utility of studying learning and variation during the first two hours of simple flaking instruction/practice, in hopes of finding a viable compromise between experimental realism and cost.

@@

*Such undertakings -- based on raw material which is never standard, and with gestures of percussion that are never perfectly delivered -- cannot be reduced to an elementary repetition of gestures...the realization of elaborate knapping activities necessitates a critical monitoring of the situation and of the decisions adopted all through the process. [@pelegrin1990: 117]*

@@

@@

## Stone Tool Making

After individual difference testing, participants engaged in a 2-hour stone tool making session, with a 15-minute break after 1 hour. Participants were instructed not to seek out additional training or information on stone tool making (i.e., via the internet) during these breaks. Each group of participants was randomly assigned to one of two experimental conditions: no teaching or teaching. In both conditions, participants were first given an opportunity to inspect and handle examples (**Fig. \@ref(fig:fig1)**) of the kind of stone tools (flakes) they are being asked to produce. They were told that their objective was to produce as many flakes as possible from the materials provided. This meant that even the untaught condition included some minimal instruction (being told the objective) , however this was considered to be unavoidable without creating a much more elaborate and naturalistic context in which participants would develop their own technological goals. Such a design would also be expected to increase behavioral variability, demanding correspondingly larger samples of participants to identify patterns and making direct comparisons with the taught condition.
```{r fig1, echo=FALSE, out.width = '75%', fig.align='center', fig.cap="Subjects examining demonstration flakes prior to the experiment. The demonstration flakes were made from the same basalt as used in the experiment with the same knapping technique."}
knitr::include_graphics(here::here("figure/Fig1.PNG"))
```

### Raw Materials

Each participant was provided with 9 cores for use over the 2-hour experiment. These cores were produced from larger chunks of a fine-grained basalt purchased from neolithics.com by fracturing them with a sledgehammer. This produced irregular, angular chunks for use in the experiment, weighing between 459g - 1876g (mean = 975g). All cores were weighed, measured (Length, Width, Thickness), and painted white so that new fracture surfaces could be discriminated from those created during production. Cores were sorted by shape and weight and then distributed evenly to each participant. As a result, there were no significant difference across participants in the mean weight (ANOVA, df = 22, F=0.3, p = 0.9; Levene test of homogeneity of variance = 1.04, df1=22, df2 = 184, p = 0.4) or shape (Length $\times$ Width/Thickness: ANOVA, df = 22, F=0.4, p = 0.9; Levene statistic = .6, df1=22, df2 = 184, p = 0.9) of cores provided. This was also true comparing the two experimental conditions (Taught vs. Untaught mean weight = 1001g vs. 956g, t = 1.24, df = 205, p = 0.2, Levene's Test F = 0.6, p = 0.4; mean shape = 221.43 vs. 221.45, t = -0.003, df = 205, p = 0.9, Levene's Test F = 3.8, p = 0.05). Participants were, however, allowed to choose which cores to work on so that differences in the weight and shape of cores actually used across participants and conditions could still emerge as a result of selection bias.

Sixty pounds of 3-to-5 inch basalt "Mexican Beach Pebbles" were purchased from a landscaping supply company for use as hammerstones in the experiment. Of these, 90 were selected as suitable for use. These weighed between 213g-1360g (mean = 425) and varied in elongation (L/W = 1.01 to 2.65) and relative thickness (LxW/T = 90.48 to 283.67). Forty-five stones were placed in the middle of the knapping area (Figure 2) for participants to freely choose from during the experiment. Broken hammerstones were replaced from the reserve to maintain a consistent number and range of choices. Each hammerstone was numbered and participants' choices were recorded along with the number of the core(s) being worked on with a particular hammerstone.

### Experimental Conditions

In both conditions, three researchers were present to record activities and collect materials. Participants were seated in a circle [@shea2015] and experiments were video recorded using two cameras (Figure 1). Participants were free to select hammerstones from the common pile and to work on any or all of their nine assigned cores in any order they preferred. However, each core and all associated debitage were collected before participants were allowed to start working on a new core, so it was not possible to partially work and then return to a particular core later. The order of cores used and associated hammerstones were recorded for each participant during the experiment.

@@

## Lithic Analysis

All finished cores were weighed and measured (L, W, T). Delta weight was calculated as (Start weight-End weight)/Start weight. All detached pieces (DPs) were collected and weighed. We did not sort DPs into types (e.g., whole flakes, fragments) as this would have greatly increased processing time and it is not clear that such distinctions add relevant information regarding utility/desirability beyond that supplied by metrics [@stout2019]. All DPs larger than 40mm in maximum dimension were photographed and measured. It is conventional in Early Stone Age lithic analysis to employ a 20 mm cut-off. We selected a higher threshold for both pragmatic (analysis time) and theoretical reasons. Flake use experiments have shown that flakes weighing less than 5--10 g or with a surface area below 7--10 cm2 [@prasciunas2007] or with a maximum dimension \<50-60 mm [@key2014] become markedly inefficient for basic cutting tasks. Similarly, data from Oldowan replication experiments [@stout2019] show that the utility index (flake cutting edge/flake mass1/3) \* (1 - exp[-0.31 \* (flake maximum dimension -- 1.81)]) developed by Morgan et al. (2015) falls off rapidly below 40mm maximum dimension ( Mean Utility \< 40mm = 0.508; \>=40mm = 0.946; t= 11.99, df = 707, p \< 0.000). By including weight in our cut-off criteria we also avoid skewing the flake shape distribution by selectively retaining long, thin pieces (i.e., MD \> 40, weight \< 5g) while discarding rounder pieces of similar (or greater) weight and area.

@@

## Statistical Analyses

To evaluate the association between psychometric, motor-skill, and training measures and technological outcomes, we adopted an information-theoretic approach [@burnham2002]. Information-theoretic approaches provide methods for model selection using all possible combinations of variables while avoiding problems associated with significance-threshold stepwise selection. We used the corrected Akaike information criterion (AICc) to rate each possible combination of predictors on the balance between goodness of fit (likelihood of the data given the model) and parsimony (number of parameters). The AICc consists of the log likelihood (i.e., how well does the model fit the data?) and a penalty term for the number of parameters that must be estimated in the model (i.e., how parsimonious is the model?), with a correction for small sample sizes (AICc converges to the standard AIC at large samples). A lower AICc indicates a more generalizable model and we used it to compare and rank various possible models. Each analysis begins with a full model that includes all predictors of interest. All possible combinations of predictors are then fit, and the resulting models are ranked and weighted based on their AICc. The "best" model is chosen because it has the lowest AICc score.

@@

# Results

Following a recent protocol to enhance the reproducibility and data transparency of archaeological research [@marwick2017], detailed results of all analyses and assessments of the data structure are available in our paper's supplementary materials and through Github (<https://github.com/Raylc/PaST-pilot>). Here we limit discussion to the major ﬁndings regarding flaking performance and individual differences. We were particularly interested in: 1) group level effects of experimental condition, 2) individual differences in aptitude and learning, and 3) potential interactions between learning conditions and individual differences. To address these questions, we employed data reduction (Principal Component Analysis) to derive two summary metrics of flaking performance, compared these factors across the two experimental conditions, and built multivariate models examining the relations between our various psychometric measures, subject's motor skill scores, and our two lithic performance factors.

## Principal Component analyses

The following two sections outline factor analyses designed to summarize our main study metrics tracking individual variation in DP sizes and shapes and lithic performance measures.

### Detached Piece size and shape

To better understand the relationship between DP shape and training/individual variation, we entered our nine flake linear plan measurements along with maximum flake length and thickness into a principal component analysis (PCA) from which summary coordinates were extracted. Bartlett's Test of Sphericity was significant ($\chi ^2$ (10) =4480, p \< .01) indicating that the set of variables are adequately related for factor analysis.

@@

### Lithic flaking performance measures

To better understand the relationship between our various lithic performance measurements and to reduce data dimensionality, we conducted a second principal component analysis examining the study's six lithic performance measures (count of large pieces [\>40mm and 5g], mass of large pieces relative to total detached mass, core delta mass, and the three flake shape factors). All of these measures were summarized for each core and unique factor scores were calculated from these core-specific measures. Bartlett's Test of Sphericity was significant ($\chi ^2$ (6) =3185, p \< .01) indicating that the set of variables are at least adequately related for factor analysis.

The analysis yielded two factors explaining a total of 56% of the variance for the entire set of variables (Table). Factor 1 (hereafter "Quantity") explains 28.7% of the variance and tracks flaking quantity due to high positive loadings on large DP count and mass ratio and on core delta mass. Performance factor 2 (hereafter "Quality") covers 27% of the sample variance and measures flaking quality as reflected in high positive loadings on Shape Factors 1 (size) and 3 (thin, "flake-like" shape) and a negative loading on Shape Factor 2 ("splinter-like" thickness and elongation). High scores on Quality thus reflect production of larger, relatively thinner, and more typically flake-shaped vs. splinter-shaped DPs.

@@

## Relationships between Performance Measures

This approach also allowed us to compare the relationship between Total Productivity, Quantity, and Quality across our two experimental groups (Figures). As might be expected, we found that per-core Quantity and Total Productivity are positively correlated in both groups (Fig. Xa), although this relationship is twice as strong in the trained (F[1, 9] = 33, p \< 0.01, Adj. $R^2$ = 0.8) compared to untrained (F[1, 8] = 8, p = 0.02, Adj. $R^2$ = 0.4) group. Interestingly, we also found evidence of a negative correlation between Total Productivity and Quality in the untrained group (F[1, 8] = 28, p = \<0.01, Adj. $R^2$ = 0.7), but no relation in the Trained group (Fig. Xb). A qualitatively similar trend with respect to Quality vs. Quantity (Fig. Xc) did not achieve significance (F[1, 17] = 0.6, p = 0.2).

@@

## Do trained, untrained, and expert knappers perform differently?

Here we compare our flaking outcomes (DP size/shape and flaking performance factors) between the trained and untrained groups. Our expert demonstrator/instructor is included as a performance benchmark.

Table summarizes the results of ANOVA tests group level difference in central tendency on various performance measures. We found no significant differences between the trained and untrained groups on our flaking Quantity and Quality factors. In contrast, three-way flake size and shape comparisons between our expert knapper and the two novice groups showed that the expert knapper made significantly more large flakes (effect size = 0.14), had a significantly higher core delta mass signal than either of the novice groups (effect size = 0.26), and left significantly smaller finished cores (effect size = 0.27) (Figure). All three of these results show either medium or large effect sizes. In all three comparisons, the trained group's data distributions tended towards the expert sample although they were not significantly different from the untrained group (Figures-Core examples too). We also observed a significant difference in shape factor 2 (splinters) driven by the expert's lower values, but with a very low (\<0.01) effect size. These results show that mean core reduction intensity and large flake production rates distinguish expert and novice performance whereas novices in experimental groups produced pieces of similar mean size and shape as those of the expert trainer.

@@

## Does performance change over time?

In addition to comparing overall performance during the two hour experiment, we also wanted to determine if groups or individuals differed in learning (i.e., performance change) over the period. For these analyses, we calculated the learning stage as the ordinal number of each core out of the total number knapped by each subject (i.e., core 2 of 4 or 4 of 8 both equal 50% complete). These relative core use-order percentages were then binned into 20 percent brackets for core-order and group-level comparisons. Flaking outcomes were tracked using the two performance factors (Quality and Quantity). We added the nodule starting mass to track whether training/practice times impacted raw material selection.

@@

## Do individual differences in motor skill and psychometric measures predict flaking performance?

One of the experiment's primary goals was to test if measures of individual perceptual-motor and cognitive variation predict success in stone flaking across different training conditions. To address this goal, we built three multivariate models examining the relations between training conditions, individual difference measures, and our three lithic performance measures (overall productivity and average per-core Quantity and Quality). These models enabled us to determine which of the psychometric and motor skill factors are better predictors of a participant's flaking performance in the study.

We considered all possible interactions between five individual difference measures, core size, training condition, and the three performance measures, with each subject providing one data point. Each model's continuous predictors (highest n-back level, Raven's Progressive Matrix score, BEAST score, starting nodule mass, Fitts score, and grip strength) were centered such that zero represents the sample average, and units are standard deviations. Our two motor skill and strength measures (grip strength and Fitt's performance scores) are also strongly correlated (F [1,19] = 15, p \< 0.01, $R^2$ = 0.41). However, these two measures track complementary components of athleticism (strength vs. speed/accuracy tradeoffs) and so we decided to include both in the model selection process.

The full models were fitted with the lm function in R 3.2.3, and we used the Glmulti package's automated model selection algorithm to select the best performing model (lowest AICc score) (see methods for further details on the multimodal selection process). All three models follow the same complete model statement as follows:

*Flaking performance variable* $\sim$ Training condition + Highest n-back level + Raven's Progressive Matrix score + BEAST score + Fitt's score + Grip strength

For our two per-core performance factors (Quantity and Quality) it is also relevant to consider how individual core features may have affected performance. We found no evidence of individual or group level practice effects over the two hours, so we did not include core order in the models. We did, however, find that subjects selected progressively larger nodules throughout the experiment. It is thus important to understand whether nodule variability had any impact on our flaking results. Because starting nodule size (mass) and shape were strongly correlated (F [1,157] = 186, p \< 0.01, $R^2$ = 0.54) we included nodule mass as a covariate to control for any variance in flaking performance that may be driven by nodule differences.

### Model 1: Individual differences and quantity flaking

Our first model examined variance in overall flaking productivity measured by each subject's combined flaked mass (nodule starting mass-core final mass). This provides a basic measure of variation in individuals' success detaching pieces and reducing cores from a standardized (see Methods) raw material supply. From the same candidate pool size of 55893 possible multivariate models, the best performing model returned an AICc value of -18 (Average AIC = -13). This model comprised the following statement with two main and three interaction effects:

*Total flaked mass* $\sim$ Training condition + Grip strength + RPM$\times$Highest n-back level + Fitts score$\times$BEAST score + Grip strength$\times$RPM

This model explains a statistically significant and substantial proportion of variance in flaking productivity ($R^2$ = 0.84, F (6, 14) = 12.7, p \< 0.01, adj. $R^2$ = 0.77). A model residuals normality test shows no significant differences with the normal distribution (p = 0.72) indicating that this relationship is linear. A Breusch-Pagan test showed no evidence for heteroskedasticity (BP = 2, df = 6, p = 0.8).

Table presents this model's coefficients and summary outputs, wherein baseline refers to the untrained condition with all continuous predictors at the sample average. The parameter estimates for the continuous predictors reflect the expected change in utility for 1 standard deviation change in the predictor variable. We found significant (p\< 0.05) and substantial (Standardized Estimate \>= 0.50, i.e. a 50% change in variable) main effects of Grip Strength, Visuospatial nBack, and BEAST. The main effect of Grip Strength (Figure), irrespective of learning condition, indicates the basic importance of strength in generating higher production rates among naive knappers at least when efficiency and quality are not considered.

@@

### Model 2: Individual differences and quality flaking

The second full model examined the variance in average flaking Quantity per core. It thus complements our first model assessing overall productivity by testing for differences in reduction intensity at the level of individual cores. From a candidate pool of 55893 possible multivariate models, the best performing model returned an AICc value of 32 (Average AIC = 44). This model comprised the following statement with three main and four interaction effects:
